# Data manipulation using `dplyr`

Trust me, this is the part of my research where I spend a significant portion of my time. Real-life data are not polished and nicely annotated. Moreover, when you want to integrate data from different sources, the fun begins (I am showing the quotation finger, of course)! Moreover, you need to format the output from one process and make it worthy for the next one. So, there's no escape from data formatting / manipulating data in real-life.

Here, we will be using the `dplyr` package which is one of the most powerful and popular packages in *R*. The `d` here represents data and `plyr` is supposed to be the tool plier. Therefore, `dplyr` packages refers to a tool to manipulate data frame. `dplyr` provides a grammar of data manipulation and the functions it provides are regarded as the verbs in the code and are very efficient ones in solving most common data manipulation problems. It is sometimes arguably more efficient than the base *R* operations.

## Install

There are mainly two ways to install `dplyr` package in **R**. You can install the `tidyverse` package and `dplyr`, being a part of it, will automatically be installed in your **R** environment. 

```{r install_dplyr_option1}
# install.packages("tidyverse")
library(tidyverse)
```

Or, you can install just the `dplyr` package by -

```{r install_dplyr_option2}
# install.packages("dplyr")
library(dplyr)
```

However, if you want to install the development version, which I won't recommend, you can follow the codes below -
```{r install_dplyr_option3}
# if (packageVersion("devtools") < 1.6) {
#   install.packages("devtools")
# }
# devtools::install_github("hadley/lazyeval")
# devtools::install_github("hadley/dplyr")

```

## Load `dplyr`

```{r load_dplyr}
library(dplyr)
```


## Pipe operator %>% 

It will be a crime not to introduce the pipe operator `%>%` to you before starting with `dplyr` verbs. If you are familiar with the pipe operator `|` in bash scripting, that's it. I have no better way to describe it to you. But, if you are not, then here is the thing for you -

The pipe operator `%>%` connects two operations on the same data (be it a vector or a data frame). It passes the output from the left-hand side operation of the pipe operator as the first argument to the right-hand side operation of the operator. If you want an informal definition -
`x %>% f(y)` is converted into `f(x,y)` by using the pipe operator. Let's look at another example. If we have a vector x that holds value from 1 to 100 and we want to calculate the mean of x and make it round to an integer, we write in base **R** -
```{r pipe_operator1}
x <- 1:100
round(mean(x))
```
However, using the pipe operator, we can first define the x and then calculate the mean and, at the end, round it to an integer, like -

```{r pipe_operator2}
x <- 1:100
x %>% mean %>% round
```

It goes from left to right as we think and build our data analysis pipeline. The new version of `dplyr` also supports `|>` as the pipe operator, but I will stick to `%>%` in the workshop. 


## dplyr verbs

There are many verbs embedded in the dplyr package, here I will be discussing a few, but very important ones, that you will need to resolve most of the data manipulation challenges in your day-to-day life. 

### select()

`select()` picks variables based on their names or types. For example -

```{r select1}
library(kableExtra)
# using specific variable names -
iris %>% select(Sepal.Length, Sepal.Width) %>% kable(align = "lccrr", caption = "iris data: Sepal length and width") %>% kable_styling(full_width = F, bootstrap_options = c("striped", "hover", "condensed")) %>% scroll_box(width = "100%", height = "200px")

# using type -
iris %>% select(is.numeric) %>% 
  kable(align = "lccrr",caption = "iris data: neumeric columns only") %>% 
  kable_styling(full_width = F, bootstrap_options = c("striped", "hover", "condensed")) %>%
  scroll_box(width = "100%", height = "200px")
```


With the verb `select()`, comes some *selection helpers* -

If you want to select all the variables, you can use `everything()`

```{r select2}
iris %>% select(everything()) %>% 
  kable(align = "lccrr",caption = "iris data: everything") %>% 
  kable_styling(full_width = F, bootstrap_options = c("striped", "hover", "condensed")) %>%
  scroll_box(width = "100%", height = "200px")
```
You can choose the last column using `last_col()` or only columns that are grouped using `group_cols()` (You will understand better when I  discuss the `group_by()` verb later).

```{r select3}
iris %>% select(last_col()) %>% 
  kable(align = "lccrr",caption = "iris data: last_col()") %>% 
  kable_styling(full_width = F, bootstrap_options = c("striped", "hover", "condensed")) %>%
  scroll_box(width = "100%", height = "200px")
iris %>% group_by(Sepal.Length,Sepal.Width) %>% select(group_cols()) %>% 
  kable(align = "lccrr",caption = "iris data: select grouped columns") %>% 
  kable_styling(full_width = F, bootstrap_options = c("striped", "hover", "condensed")) %>%
  scroll_box(width = "100%", height = "200px")
```

If there's a common prefix or suffix to some column names, you can utilise that by using selectio helpers `starts_with()` or `ends_with()`, respectively -

```{r select4}
iris %>% select(starts_with("Sepal")) %>% 
  kable(align = "lccrr",caption = "iris data: columns starts with Sepal") %>% 
  kable_styling(full_width = F, bootstrap_options = c("striped", "hover", "condensed")) %>%
  scroll_box(width = "100%", height = "200px")

iris %>% select(ends_with("Length")) %>% 
  kable(align = "lccrr",caption = "iris data: columns ends with Length") %>% 
  kable_styling(full_width = F, bootstrap_options = c("striped", "hover", "condensed")) %>%
  scroll_box(width = "100%", height = "200px")
```
Even, an internal pattern can be used to select a column by using `contains()` -
```{r select5}
iris %>% select(contains("dth")) %>% 
  kable(align = "lccrr",caption = "iris data: column names containing 'dth'") %>% 
  kable_styling(full_width = F, bootstrap_options = c("striped", "hover", "condensed")) %>%
  scroll_box(width = "100%", height = "200px")
```

Even, you can use regular expression to select a column by using `matches()` -
```{r select6}
# column name containing either W or d or both
iris %>% select(matches("[Wd]")) %>% 
  kable(align = "lccrr",caption = "iris data: column name containing W or d") %>% 
  kable_styling(full_width = F, bootstrap_options = c("striped", "hover", "condensed")) %>%
  scroll_box(width = "100%", height = "200px")
```


### filter()

The `filter()` verb is used to subset a data frame based on one or more conditions. Only the rows that satisfy the condition(s) remain and others filter out. There are some functions and operators that you should know while dealing with `filter()` verb -

`==`, `>` or `<`, `>=` or `<=`
`&`, `|`,  `!`
`is.na()`
`%in%`

Let's see some examples -
```{r filter}

# choose the rows whose Petal.Width is greater than 2
iris %>% filter(Petal.Width > 2) %>% 
  kable(align = "lccrr",caption = "iris data: Petal width creater than 2") %>% 
  kable_styling(full_width = F, bootstrap_options = c("striped", "hover", "condensed")) %>%
  scroll_box(width = "100%", height = "200px")

# choose the rows for setosa Species

iris %>% filter(Species == "setosa") %>% 
  kable(align = "lccrr",caption = "iris data: setosa only") %>% 
  kable_styling(full_width = F, bootstrap_options = c("striped", "hover", "condensed")) %>%
  scroll_box(width = "100%", height = "200px")
#or 
iris %>% filter(Species %in% "setosa") %>% 
  kable(align = "lccrr",caption = "iris data: setosa only") %>% 
  kable_styling(full_width = F, bootstrap_options = c("striped", "hover", "condensed")) %>%
  scroll_box(width = "100%", height = "200px")

# or even the opposite is True
iris %>% filter(Species != "setosa") %>% 
  kable(align = "lccrr",caption = "iris data: without setosa") %>% 
  kable_styling(full_width = F, bootstrap_options = c("striped", "hover", "condensed")) %>%
  scroll_box(width = "100%", height = "200px")
```


### mutate()

The verb `mutate()` creates new columns and often the element of the new col can be functions of the existing variables (i.e. columns).

```{r mutate}
iris %>% mutate(Length_difference = Sepal.Length - Petal.Length) %>% # not that the new column here make much sense
  kable(align = "lccrr",caption = "iris data: new column added") %>% 
  kable_styling(full_width = F, bootstrap_options = c("striped", "hover", "condensed")) %>%
  scroll_box(width = "100%", height = "200px")

# To keep only the newly created column, use transmute()
iris %>% transmute(Length_difference = Sepal.Length - Petal.Length) %>% 
  kable(align = "lccrr",caption = "iris data: new column only") %>% 
  kable_styling(full_width = F, bootstrap_options = c("striped", "hover", "condensed")) %>%
  scroll_box(width = "100%", height = "200px")
```

Interestingly, setting the value of an existing column to `NULL` inside mutate deletes the column.



### rename()

As the name suggests, `rename()` verb changes the name of an existing column. The syntax is `<new_name> = <old_name>`. Example -
```{r rename1}
iris %>% rename(Species.name=Species) %>% 
  kable(align = "lccrr",caption = "iris data: Species column renamed") %>% 
  kable_styling(full_width = F, bootstrap_options = c("striped", "hover", "condensed")) %>%
  scroll_box(width = "100%", height = "200px")
```
Interestingly, you can change the name of a column while selecting using `select()` verb -

```{r rename2}
iris %>% select(Sepal.Length, 
                Sepal.Width, 
                Petal.Length, 
                Petal.Width, 
                Species.name=Species) %>% 
  kable(align = "lccrr",caption = "iris data: Species column renamed using select()") %>% 
  kable_styling(full_width = F, bootstrap_options = c("striped", "hover", "condensed")) %>%
  scroll_box(width = "100%", height = "200px")
```


### arrange()

The verb `arrange()` arranges or orders the rows of a data frame by the values of selected column(s), like -

```{r arrange}
iris %>% arrange(Sepal.Length) %>% 
  kable(align = "lccrr",caption = "iris data: arranged by Sepal length") %>% 
  kable_styling(full_width = F, bootstrap_options = c("striped", "hover", "condensed")) %>%
  scroll_box(width = "100%", height = "200px")

# After arranging the data frame by Sepal.Length, for a distinct Sepal.Length, the Sepal.Width is arrange and so as the rest of the data frame with it.
iris %>% arrange(Sepal.Length,Sepal.Width) %>% 
  kable(align = "lccrr",caption = "iris data: arranged by Sepal length and width") %>% 
  kable_styling(full_width = F, bootstrap_options = c("striped", "hover", "condensed")) %>%
  scroll_box(width = "100%", height = "200px")
```


### distinct()

The `distinct()` verb retains only the unique/distinct rows from a data frame given the column(s) selected and only output the select column(s), if not the `.keep_all` parameter is change from it's default value `FALSE` to `TRUE`. Let's see some examples -

```{r distinct}
iris %>% distinct(Sepal.Length) %>% 
  kable(align = "lccrr",caption = "iris data: distinct Sepal length") %>% 
  kable_styling(full_width = F, bootstrap_options = c("striped", "hover", "condensed")) %>%
  scroll_box(width = "100%", height = "200px")

# here only the unique combinations of Sepal.Length and Sepal.Width are kept.
iris %>% distinct(Sepal.Length,Sepal.Width) %>% 
  kable(align = "lccrr",caption = "iris data: distinct Sepal length and width only") %>% 
  kable_styling(full_width = F, bootstrap_options = c("striped", "hover", "condensed")) %>%
  scroll_box(width = "100%", height = "200px")


# rest of the columns are also returned.
iris %>% distinct(Sepal.Length,Sepal.Width, .keep_all = T) %>% 
  kable(align = "lccrr",caption = "iris data: distinct Sepal length and width only") %>% 
  kable_styling(full_width = F, bootstrap_options = c("striped", "hover", "condensed")) %>%
  scroll_box(width = "100%", height = "200px")

```



### slice()

the `slice()` verb lets you index rows by their (integer) locations. It has some helpers too -


`slice_head()` selects the first row, while `slice_tail()` selects the last. Same can be done using `slice(1)` and `slice(n())`

`slice_head(<int>)` selects from the first to the `<int>th` row, while `slice_tail(<int>)` selects from `<int>th` to the last row up to the end row.

`slice_sample()` selects rows at random

`slice_min()` and `slice_max()` helper selects rows with the lowest and the highest value of the selected variable. Few examples -

```{r slice}
iris %>% slice(1) %>% 
  kable(align = "lccrr",caption = "iris data: a random row") %>% 
  kable_styling(full_width = F, bootstrap_options = c("striped", "hover", "condensed")) %>%
  scroll_box(width = "100%", height = "200px")
iris %>% slice(10:n()) %>% 
  kable(align = "lccrr",caption = "iris data: from 10th row to the end") %>% 
  kable_styling(full_width = F, bootstrap_options = c("striped", "hover", "condensed")) %>%
  scroll_box(width = "100%", height = "200px")

iris %>% slice_min( Sepal.Length) %>% 
  kable(align = "lccrr",caption = "iris data: row with the lowest sepal length") %>% 
  kable_styling(full_width = F, bootstrap_options = c("striped", "hover", "condensed")) %>%
  scroll_box(width = "100%", height = "200px")
```



### join

A disclaimer, there's no verb (exactly) called `join()` in dplyr (at least, to date). However, there are two types of join verbs - `inner_join()` and `outer_join` (which is also not a verb, but a class of three verbs: `left_join()`, `right_join()` and `full_join()`). join verbs joins columns from two data frames based on a common key column.

`inner_join()` verb joins two data frame and retains the rows where the keys match. This means that there is a potential loss of observations that we may not appreciate in the real-life analysis.

On the other hand, if we have two data frames `x` and `y`, the `left_join()` verb matches the keys from `x` and `y`, while keeps all the rows from `x` and joins the matched rows from `y`. The empty cells are filled with NA values. For `right_join()` verb, is the opposite scenario. On the other hand, the `full_join()` verb retains all the rows from both data frames and empty cells are filled with NA values. Let's clear the concept with some examples -
```{r join}
x <- iris %>% select(Sepal.Length,Sepal.Width,Species) %>% filter(Species %in% c("setosa", "versicolor")) %>% slice_sample(n=10)

y <- iris %>% select(Petal.Length,Petal.Width,Species) %>% filter(Species %in% c("versicolor", "virginica")) %>% slice_sample(n=10)


x %>% inner_join(y, by = "Species") %>% 
  kable(align = "lccrr",caption = "iris data: inner_join") %>% 
  kable_styling(full_width = F, bootstrap_options = c("striped", "hover", "condensed")) %>%
  scroll_box(width = "100%", height = "200px")

x %>% left_join(y, by = "Species") %>% 
  kable(align = "lccrr",caption = "iris data: left_join") %>% 
  kable_styling(full_width = F, bootstrap_options = c("striped", "hover", "condensed")) %>%
  scroll_box(width = "100%", height = "200px")

x %>% right_join(y, by = "Species") %>% 
  kable(align = "lccrr",caption = "iris data: right_join") %>% 
  kable_styling(full_width = F, bootstrap_options = c("striped", "hover", "condensed")) %>%
  scroll_box(width = "100%", height = "200px")

x %>% full_join(y, by = "Species") %>% 
  kable(align = "lccrr",caption = "iris data: full_join") %>% 
  kable_styling(full_width = F, bootstrap_options = c("striped", "hover", "condensed")) %>%
  scroll_box(width = "100%", height = "200px")

```



### group_by() and summarise()

I will be describing `group_by()` and `summarise()` verbs together to show the effect of the former. `group_by()` is the most importsnt grouping verb in dplyr. It takes one or more variables of the data frame to group by -

```{r group_by1}
iris %>% group_by(Species) %>% 
  kable(align = "lccrr",caption = "iris data: group_by Species") %>% 
  kable_styling(full_width = F, bootstrap_options = c("striped", "hover", "condensed")) %>%
  scroll_box(width = "100%", height = "200px")
```

Rather than some metadata, you don't see any change in the structure of the iris data frame yet. Let's select Sepal.Length and see the effect -

```{r group_by2}
iris %>% group_by(Species) %>% select(Sepal.Length) %>% 
  kable(align = "lccrr",caption = "iris data: group by Species and selected by Sepal length") %>% 
  kable_styling(full_width = F, bootstrap_options = c("striped", "hover", "condensed")) %>%
  scroll_box(width = "100%", height = "200px")
```

Though I selected only the `Sepal.Length`, the Species column also appears. Yes, that's because of the application `group_by()` verb beforehand. But the most dramatic effect can be seen in conjunction with the `summarise()` verb. 

`summarise()` generates a new data frame and returns one row (with the result of course) for each combination of grouping variables. In the case of no grouping variables, the output has a single row summarising all observations in the input. Now, let's see the effect of `group_by()` in conjunction with `summarise()` verb -

```{r summarise}
iris %>% 
  group_by(Species) %>% 
  select(Sepal.Length) %>% 
  summarise(count=n()) %>% 
  kable(align = "lccrr",caption = "iris data: summarised count by Species") %>% 
  kable_styling(full_width = F, bootstrap_options = c("striped", "hover", "condensed")) %>%
  scroll_box(width = "100%", height = "200px")

iris %>% 
  group_by(Species) %>% 
  select(Sepal.Length) %>% 
  summarise(mean_Sepal_length=mean(Sepal.Length)) %>% 
  kable(align = "lccrr",caption = "iris data: Summarised mean Sepal length by Species") %>% 
  kable_styling(full_width = F, bootstrap_options = c("striped", "hover", "condensed")) %>%
  scroll_box(width = "100%", height = "200px")

# as being told, without any grouping -

iris %>% 
  select(Sepal.Length) %>% 
  summarise(mean_Sepal_length=mean(Sepal.Length)) %>% 
  kable(align = "lccrr",caption = "iris data: summarised mean Sepal length without grouping") %>% 
  kable_styling(full_width = F, bootstrap_options = c("striped", "hover", "condensed")) %>%
  scroll_box(width = "100%", height = "200px")

```





